---
title: "EssentiallyR: Data Management"
output: html_notebook
---

This notebooks is a collection of notes from my studies from [EssentilR](https://onlinecourses.science.psu.edu/stat485/sites/onlinecourses.science.psu.edu.stat485/files/EssentialR/index.pdf). The next couple of sections within this notebook cover the following topics:
* R as a DataFrame
* Importing and Manipulating Data
* Working with multiple variables

### The DataFrame

```{r}
a <- list(c(1,2,3), "Blue", factor(c("A", "B","A", "B","B")))
a
```
`data()` is a function to load in "built in" data sets are stored as data frame.

```{r}
cyl<-factor(scan(text= "6 6 4 6 8 6 8 4 4 6 6 8 8 8 8 8 8 4 4 4 4 8 8 8 8 4 4 4 8 6 8 4"))
am<-factor(scan(text= "1 1 1 0 0 0 0 0 0 0 0 0 0 0 0 0 0 1 1 1 0 0 0 0 0 1 1 1 1 1 1 1"))
levels(am) <- c("auto", "manual")

disp<-scan(text= "2.62 2.62 1.77 4.23 5.90 3.69 5.90 2.40 2.31 2.75 2.75 4.52 4.52 4.52
7.73 7.54 7.21 1.29 1.24 1.17 1.97 5.21 4.98 5.74 6.55 1.29 1.97 1.56 5.75 2.38 4.93 1.98")
hp<-scan(text="110 110 93 110 175 105 245 62 95 123 123 180 180 180 205 215 230 66
52 65 97 150 150 245 175 66 91 113 264 175 335 109")

# create data frame for the cars dataset

car <- data.frame(cyl, disp, hp, am)
head(car)

```


```{r}
summary(car)
names(car)
```

View and modify column headings
```{r}
names(car)
names(car)[4]
names(car)[4] <- "trans"
names(car)
car$trans
```
Using the `[]` _index operator_ to access parts of the dataframe by _rows & columsns_. Leaving one value black implies "all rows" or "all columns". 

```{r}
# this gives us rows 1:3
car[1:3, ]

```


#### Subset
`subset()` function which is useful for workng with dataframes, since it allows you to extract data from the dataframe based on multiple conditions, and it has n easy to read syntax. 
```{r}
data(faithful)
summary(subset(faithful, eruptions <= 3))
#plot(faithful)
```

### Chaning Data Frames (adding new column)

Its very simple to add new columns (or rows) to a df 
```{r}
# specify a new name for a variable can create a new column
car[,5] <- as.numeric(levels(car$cyl)[car$cyl])
names(car)
names(car)[5] <- "cyl.numeric"
# data set now has 5 columns 'V5' - 'Variable 5'

#lets calculate displacement per cylinder
car$disp.per.cyl <- car$disp/car$cyl.numeric
names(car)


```

#### combining data frames together
- `cbind()` - column wise binding
- `rbind()` - row wise binding

```{r}
# combind mtcars variables that are not in the cars dataset
data(mtcars)
names(mtcars)

dim(car)
car <- cbind(car, mtcars[, c(1, 5:8, 10:11)])
dim(car)
head(car)

```

#### Comments
`comments()` this function is an attribute of data frames that is reserved for comments.
```{r}
comment(car) <- "A data set derived from the mtcars dataset. Displacement is in liters"

```

# Chapter 9: Manipulating Data (Data Wrangling)

### Summarizing Data

```{r}
setwd("~/R/essentiallyR")
beans <- read.csv("./Data/BeansData.csv", header = TRUE, comm = "#")
dim(beans)
```

`apply()` is a family of functinos used to _"do something over and over again to subset of some data"

Below:
* we apply function `mean()` to the columns (MARGIN=2) 6,7,8 of beans
* columns 3,4, and 5 are factors so means() would give an error
* `na.rm = True` refers to arguments that are passed to the function specififed by FUN `mean()`

```{r}
# get the means for columns of dta
apply(X = beans[,6:8], MARGIN = 2, FUN = mean, na.rm = TRUE)

```

Summariaze data to get __group means__: means for each treatment type:

```{r}
tapply(beans$rt.len, INDEX = list(beans$trt), FUN = mean, na.rm = TRUE)

# variable trt was coded of tall the treatments

```


```{r}
tapply(beans$rt.len, list(beans$pot.size, beans$P.lev), mean, na.rm=TRUE)
```


```{r}
tapply(beans$rt.len, list(paste(beans$pot.size, beans$P.lev)), mean , na.rm=TRUE)
```

Often we want to get summary data for multiple columns. The `aggregate()` is a convience form of apply that makes this trivially easy. 

```{r}
aggregate(x = beans[, 6:8], by = list(beans$pot.size, beans$phos), FUN = mean, na.rm = TRUE)

```


```{r}
# same as above but using column numbers instead of column names for a nicer output

aggregate(x = beans[, 6:8], by = beans[c(1,2)], FUN = mean, na.rm = TRUE)
aggregate(x = beans[, 6:8], by = beans[c(1,2)], FUN = sd, na.rm = TRUE)
```

Specify my own function for aggregation by defining `function(x)` on the fly

```{r}
aggregate(beans[, 6:8], beans[c(1,2)], function(x) (sd(x, na.rm = TRUE) / (length(x) - sum(is.na(x))) ^0.5))
```



```{r}
beans.means <- aggregate(beans[, 6:8], beans[c(1,2)], mean, na.rm = TRUE)

# re-rder the the data for nicer plotting
beans.means <- beans.means[order(beans.means$pot.size, beans.means$phos),]

#add labels and legend
barplot(beans.means$ShtDM, 
        col = c("white", "gray70"), 
        names.arg = paste(beans.means$pot.size, beans.means$phox, sep="\n"), 
        ylab = "Shoot biomass (g)")
legend("topright", fill=c("white", "gray70", "grey50"), legend=c("LP", "HP"))
```

### Reformatting Data from "Wide" to "long"

```{r}
data("PlantGrowth")
head(PlantGrowth)
```

`unstack()` used to make one column for each level of group

```{r}
unique(PlantGrowth$group)
unstack(PlantGrowth)

```



```{r}
pg <- unstack(PlantGrowth, weight ~ group)
boxplot(pg)

```

### Reshape
`reshape` package hs powerful tools to reshape data in many ways. The two core functions are `melt()` and `cast()` (thinking working with metal) - once you 'melt' the data you can 'cast' it into any shape you want. 
* `melt()` turns a data set into a series of observations which consist of a variable and value
* `cast()` reshapes melted data, 'casting' it into a new form

```{r}
require(reshape)
data(iris)
iris$id <- row.names(iris)
head(melt(iris, id = "id"))

```



```{r}
tail(melt(iris, id = "id"))
```



```{r}
melt.iris <- melt(iris, id = c("id", "Species"))
dim(melt.iris)

```
Now instead of 150 observations with 6 variables we have 600 observations with 4 variables.

```{r}
head(melt.iris)

```



```{r}
cast(melt.iris, Species ~ variable, mean)
```

`strplit()` used to split the strngs that represents variables

```{r}
# separate out Sepal and Petal variables
head(strsplit(as.character(melt.iris$variable), split = ".", fixed = TRUE))

```

Notice that this returns a list. We’ll have to use do.call() to call rbind() on the list to bind the list
elements into a data frame.

```{r}
head(do.call(rbind, strsplit(as.character(melt.iris$variable), split = ".",
fixed = TRUE)))

```


```{r}
melt.iris <- cbind(melt.iris, do.call(rbind, strsplit(as.character(melt.iris$variable), 
                                                      split = ".", fixed = TRUE)))
names(melt.iris)[5:6] <- c("Part", "Dimension")
head(melt.iris)
```


```{r}
cast(melt.iris, Species ~ Dimension | Part, mean)
```


```{r}
cast(melt.iris, Species ~ Dimension + Part, mean)
```


```{r}
cast(melt.iris, Species ~ Dimension, mean)
```


```{r}
# back to original data
head(cast(melt.iris, Species + id ~ Part + Dimension))
```

### Mearging Data Sets

```{r}
authors <- data.frame(surname = c("Tukey", "Venables", "Tierney", "Ripley","McNeil"), 
                      nationality = c("US", "Australia", "US", "UK", "Australia"),
deceased = c("yes", rep("no", 4)))
books <- data.frame(name = c("Tukey", "Venables", "Tierney", "Ripley", "Ripley","McNeil", "R Core"), title = c("Expl. Data Anal.", "Mod. Appl. Stat ...","LISP-STAT", "Spatial Stat.", "Stoch. Simu.", "Inter. Data Anal.", "An Intro. to R"),
other.author = c(NA, "Ripley", NA, NA, NA, NA, "Venables & Smith"))
authors

```


```{r}
# merge the 2 data frames together
(m1 <- merge(authors, books, by.x = "surname", by.y = "name"))
```


```{r}
(m2 <- merge(books, authors, by.x = "name", by.y = "surname"))
```

###  More about Loops
```{r}
for (i in 1:5) print(i)
```


```{r}
x <- c(2, 5, 6, 23, 6)
for (i in x){
  cat(paste("i^2=", i^2, "\n"))
}
```


```{r}
par(mfrow=c(3,1),mar=c(3,3,0.5,0.5))
for (p in c(4,5,3)){
barplot(beans.means[,p],col=c("white","grey70"),
names.arg=paste(beans.means$pot.size,beans.means$phos,sep="\n"))
# plot the pth column of beans.means
}

```
