---
title: "DUKE - t-distributions and comparing 2 means"
output: html_notebook
---

https://www.coursera.org/learn/inferential-statistics-intro/


finding propbabilities

P(|Z| > 2)
```{r}
pnorm(2, lower.tail = FALSE) * 2
```
P(|t.df = 50| > 2)
```{r}
pt(2, df = 50 ,lower.tail = FALSE) * 2
```
P(|t.df = 10| > 2) 

```{r}
pt(2, df= 10, lower.tail = FALSE) * 2
```


```{r}
qt(0.025, df = 21 )
```


```{r}
pt(30, df = 21, lower.tail = FALSE)
```

Estimate the difference between the average post-mean snack consumption bween those who eat with and without distractions

```{r}
x_w <- 52.1
s_w <- 45.1
n_w <- 22

x_wo <- 27.1
s_wo <- 26.4
n_wo <- 22


# point estimates
point_estimates <- x_w - x_wo

# DF for t statistic for inference on differnce of two means
dft <- min(n_w -1, n_wo -1)

# critical statisitc (t)
t_star <- abs(qt(0.025, df = dft))

# Standard Error
se <- sqrt(((s_w^2)/n_w) + ((s_wo^2)/n_wo))

confidence_interval <- t_star * se

# confidence interval
low <- point_estimates - confidence_interval
high <- point_estimates + confidence_interval

low
high

```


```{r}

# hypothesis
# Ho: mu(w) - mu(wo) = 0
# Ha: mu(w) - mu(wo) != 0

# t-value
t_value <- (point_estimates - 0) / se


# p-value
p_value <- pt(t_value, df = dft, lower.tail = FALSE) * 2

p_value
p_value < 0.05


```

### Inference for comparing two paired means

Calculate the test statistic and the p-value for this hypothesis test
https://www.coursera.org/learn/inferential-statistics-intro/lecture/k5zhM/inference-for-comparing-two-paired-means

```{r}
t_value <- abs(-0.87)
p_value <- pt(t_value, df = (200-1), lower.tail = FALSE) * 2
each_tail_area <- p_value /2
each_tail_area
p_value
p_value < 0.05
```

### Confidence Intervals and Sampling
A sample of 50 college students were asked how many exclusive relationships they've been in so far. The students in the sample had an average of 3.2 exlusive relathipships with a standard deviation of 1.74. In addition, the sample distribution was only slightly skewed to the right. Estimate the true average number of exclusive relationships based on this sample using a 95% confidence interval.

https://www.coursera.org/learn/inferential-statistics-intro/lecture/txXVo/ci-for-the-mean-examples

```{r}
n <- 50
x_bar <- 3.2
s <- 1.74
ci <- 0.95
alpha <- 1-ci

# standard error
se <- (s/sqrt(n))

# z score (CRITICAL VALUE)
z <- abs(qnorm((alpha)/2))

# Confidence Intervals
ci_low <- x_bar - (z*se)
ci_high <- x_bar + (z*se)

ci_low
ci_high
```

We are 95% confident that college students on average have been in 2.75 to 3.68 exclusive relationships

### p-value
p(xbar > 3.2 | ho: mu = 3)
```{r}
mu <- 3

#z-score (TEST STATISTIC)
z_score <- (3.2-3)/se

# p value
p_value <- pnorm(-abs(z_score))
p_value
p_value < alpha

```
p-value = P(Z > 0.81) = 0.2090.20
Our p-value is 0.209, and since it's high, at least, it's higher than 5%, we do not reject the null hypothesis.


```{r}
n <- 36
x_bar <- 118.2
s <- 6.5
ci <- 0.99
mu <- 100
alpha <- 1-ci

# standard error
se <- (s/sqrt(n))

# z score (CRITICAL VALUE)
z <- abs(qnorm((alpha)/2))

# Confidence Intervals
ci_low <- x_bar - (z*se)
ci_high <- x_bar + (z*se)

ci_low
ci_high


#z-score (TEST STATISTIC)
z_score <- (x_bar - mu)/se

# p value
p_value <- pnorm(-abs(z_score))
p_value
p_value < alpha
```
